Approximation as Differentiation

### 2025-05-24 main/formal-structures.tex
Slow prog..

$\LatGal$ is not Cartesian closed, and so we cannot use it directly to interpret higher-order programs. In
fact Cartesian closure is incompatible with another notable property of $\LatGal$, which follows from the fact
that the projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y \to Y$, where $X \times Y$ denotes the
product of lattices, have both upper and lower adjoints. This allows $X \times Y$, which we shall hereafter
write as $X \biprod Y$, to act as both a product and a coproduct, with projections $\biproj_X$ and $\biproj_Y$
and injections $\biinj_X$ and $\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and $\cat{C}$ has only a single object, up to isomorphism. Although the biproduct
structure rules out Cartesian closure, it will prove crucial for establishing a Cartesian closed setting in
\secref{higher-order}; first we develop a suitable categorical setting for first-order programs.

>>>>

$\LatGal$ is not Cartesian closed, and so cannot serve directly as the basis for interpreting higher-order
programs. In fact Cartesian closure is incompatible with another notable property of $\LatGal$. The
projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y \to Y$, where $X \times Y$ denotes the product of
lattices, have both upper and lower adjoints; this means that $X \times Y$, which we shall hereafter write as
$X \biprod Y$, is both a product and a coproduct, with projections $\biproj_X$ and $\biproj_Y$ and
injections $\biinj_X$ and $\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and $\cat{C}$ has only a single object, up to isomorphism. However although the
biproduct structure rules out Cartesian closure, it will prove crucial for establishing a Cartesian closed
setting in \secref{higher-order}; but let us first see how to develop a suitable categorical setting for
first-order programs.

### 2025-05-24 main/framework.tex
Slow prog..

What we want to do now is understand how to equip the interpretation of a (first-order) program with a Galois
connections relating approximations of its inputs to approximations of its output. Since these lattices of
approximations depend on the value being approximated, the semantics of every type $\tau$ must have both a
``conventional'' component, providing the set $X$ which interprets $\tau$, plus an associated lattice of
approximations $\partial X(x)$ for every point $x \in X$. The semantics of an expression $e$ of type $\sigma$
with a single free variable of type $\tau$ (with $\tau$ and $\sigma$ denoting $X$ and $Y$) must then also have
two components: a function $f: X \to Y$ giving the conventional interpretation of $e$, and a family of Galois
connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for every $x \in X$ which can be used for
slicing over that expression.

This pattern of objects and morphisms is exactly the \emph{Grothendieck construction} for a particular indexed
category $\Set^\op \to \Cat$, namely the indexed category $\Func{-}{\LatGal}$ which assigns to every set $X$
the functor category $\Func{X}{\LatGal}$. Here the (discrete) category $X$ is a plain set and so
$\Func{X}{\LatGal}$ is exactly the category of $X$-indexed families of bounded lattices, with $X$-indexed
families of Galois connections between them as morphisms. The Grothendieck construction
$\Grothendieck{}T$~(\defref{Grothendieck} below) for an indexed category $T$, sometimes called the \emph{total
category} for $T$, is the category of all $\Set$-indexed families of objects of $T$, together with morphisms
that account for both structure within each family and how families transform under reindexing.

>>>>

The goal now is to equip the interpretation of a (first-order) program with a Galois connection relating
approximations of its inputs to approximations of its output. Since the specific lattices of approximations
depend on the value being approximated, the semantics of every type $\tau$ must have both a ``conventional''
component, providing the set $X$ which interprets $\tau$, plus an associated lattice of approximations
$\partial X(x)$ for every point $x \in X$. The semantics of an expression $e$ of type $\sigma$ with a single
free variable of type $\tau$ (with $\tau$ and $\sigma$ denoting $X$ and $Y$) must then also have two
components: a function $f: X \to Y$ giving the conventional interpretation of $e$, and a family of Galois
connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for every $x \in X$ which can be used for
slicing over that expression.

This pattern of objects and morphisms is captured by the construction used in CHAD~\cite{vákár22,nunes2023} to
describe automatic differentiation, namely the \emph{Grothendieck construction} for a particular indexed
category $T: \Set^\op \to \Cat$. For CHAD $T(X)$ is the functor category $\Func{X}{\FinVect}$, regarded as the
category of $X$-indexed families of finite vector spaces, with ; here $T(X) = \Func{X}{\LatGal}$, the category of
$X$-indexed families of bounded lattices, with $X$-indexed families of Galois connections between them as
morphisms. The Grothendieck construction $\Grothendieck{}T$~(\defref{Grothendieck} below) for an indexed
category $T$, sometimes called the \emph{total category} for $T$, is the category of all $\Set$-indexed
families of objects of $T$, together with morphisms that account for both structure within each family and how
families transform under reindexing.
